---
title: "UKB - Heteroplasmy, CHIP, and MN"
subtitle: "04 - survival analysis by subtype: Figure 6, Supplementary Figure 12"  
author: "Yun Soo Hong"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false 
      smooth_scroll: false 
    pandoc_args: [
    "--number-sections",
    "--number-offset=0"
    ]
fontsize: 11pt              
---

# Set up  

## Program description  
* This program reads in the cleaned data set (`ukb_mthet_chip_dp3_2024_03_13.rds`), performs analysis in the "Mitochondrial heteroplasmy improves risk prediction for myeloid neoplasms" paper (Nature Communications, 2024), and provided the data used for generating tables and figures.   
* This file contains
  -   Figure 6, Supplementary Figure 12.    

* Steps  
  * Step 1: Read in data  
  * Step 2: Settings    
  * Step 3: Figure 6  
  * Step 4: Supplementary Figure 12    

## Version of R and packages  

```{r analysis set up, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries  

```{r analysis load libraries, echo = TRUE, message = FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(data.table)
library(survival)
library(survminer)
```

## Print out settings  

```{r analysis print settings}
options(tibble.print_max = 100)
options(pillar.sigfig = 5)
```

# Load data  

```{r analysis load data, include = FALSE}
ukb <- 
  readRDS("../data_cleaning/ukb_mthet_chip_dp3_2024_03_13.rds")
```

# Settings   

## Excel output  

```{r analysis excel output}
wb <- openxlsx::createWorkbook("")

# Add some sheets to the workbook
openxlsx::addWorksheet(wb, "Figure 6")
openxlsx::addWorksheet(wb, "Supplementary Figure 12")
```

## Generate new variables      

```{r analysis variables}
summary(ukb$cancer_date_blood_mn)

# all participants  
ukb <-
  ukb |>
  mutate(
    # subtypes of MN  
    subtype = case_when(
      cancer_dx_unknown_mds == 1 & 
        cancer_date_unknown_mds ==
        pmin(cancer_date_unknown_mds, cancer_date_blood_leukemia_aml, cancer_date_blood_mpn, cancer_date_blood_mn_other) ~ 1,
      cancer_dx_blood_leukemia == 1 & 
        cancer_date_blood_leukemia_aml ==
        pmin(cancer_date_unknown_mds, cancer_date_blood_leukemia_aml, cancer_date_blood_mpn, cancer_date_blood_mn_other) ~ 2,
      cancer_dx_blood_mpn == 1 &
        cancer_date_blood_mpn ==
        pmin(cancer_date_unknown_mds, cancer_date_blood_leukemia_aml, cancer_date_blood_mpn, cancer_date_blood_mn_other) ~ 3,
      cancer_dx_blood_mn_other == 1 &
        cancer_date_blood_mn_other ==
        pmin(cancer_date_unknown_mds, cancer_date_blood_leukemia_aml, cancer_date_blood_mpn, cancer_date_blood_mn_other) ~ 4,
      TRUE ~ 0
    ),
    subtype = factor(
      subtype, 
      levels = c(0:4),
      labels = c("none", "mds", "aml", "mpn", "other")
    )
  ) 
```

# Figure 6 - Hazard ratios (95% confidence intervals) for the associations of CHRS-M categories with subtypes of MN.    

## Myeloid neoplasms  

```{r analysis figure 6 mn}
ukb_chip <- 
  ukb |>
  filter(chip_yn == 1) |>
  mutate(
    # time to event 
    time = as.numeric((cancer_date_blood_mn - dov)/365.25),
    end_date = case_when(
      # those who develop MN before cancer or death, among those who did not have previous cancer   
      cancer_dx_blood_mn == 1 & prev_cancer_yn == 0 & 
        (cancer_date_blood_mn <= cancer_date_cancer & cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      # those who develop MN before cancer or death, among those who had previous cancer   
      cancer_dx_blood_mn == 1 & prev_cancer_yn == 1 & 
        (cancer_date_blood_mn > cancer_date_cancer & cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      # those who develop cancer before developing MN or death 
      prev_cancer_yn == 0 & cancer_dx_cancer == 1 & 
        (cancer_date_cancer < cancer_date_blood_mn & cancer_date_cancer < date_of_death) ~ cancer_date_cancer,
      # those who die without developing MN or cancer
      death == 1 & 
        (date_of_death < cancer_date_cancer | date_of_death < cancer_date_blood_mn) ~ date_of_death, 
      # otherwise, administrative censoring  
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25, 
    
    high_risk_fct = droplevels(high_risk_fct),
    single_dnmt3a_fct = droplevels(single_dnmt3a_fct),
    mutation_n_fct = droplevels(mutation_n_fct), 
    max_vaf_fct = droplevels(max_vaf_fct), 
    chrs_cat = droplevels(chrs_cat),
    
    # MSS category, reduced
    mss_cat3 = case_when(
      mMSS < 0.51 ~ 0, 
      mMSS >= 0.51 & mMSS < 1.52 ~ 1, 
      mMSS >= 1.52 & !is.na(mMSS) ~ 2
    ),
    mss_cat3 = factor(
      mss_cat3,
      levels = c(0:2), 
      labels = c("MSS <0.51", "MSS 0.51-1.52", "MSS 1.52+")
    ),
    
    # raw scores + heteroplasmy yes/no, MSS in 3 categories for CHIP (+)    
    chrsm = (chrs + I(het_yn == 0)*1 + I(het_yn == 1)*1.5 + 
                  I(mss_cat3 == "MSS <0.51")*1 + I(mss_cat3 == "MSS 0.51-1.52")*1.5 + I(mss_cat3 == "MSS 1.52+")*2),
    chrsm = as.numeric(chrsm),
    
    # adding cutoff by 1 from original CHRS  
    chrsm_cat = case_when(
      chrsm <= 11.5 ~ 0,  
      chrsm >= 12 & chrsm <= 14 ~ 1, 
      chrsm >= 14.5 ~ 2
    ), 
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c(0:2), 
      labels = c("low", "intermediate", "high")
    )
  )

surv_chip <-
  with(ukb_chip, Surv(time = tt, event = cancer_dx_blood_mn))

# CHRS-M category
chrsm_cat_mn <- survival::coxph(surv_chip ~ chrsm_cat + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chrsm_cat_mn)

# components of CHRS-M
chip_mn <- survival::coxph(surv_chip ~ 
                             age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct + 
                             het_yn + mss_cat3 + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chip_mn)
```

## AML  

```{r analysis figure 6 aml}
ukb_chip <- 
  ukb |>
  filter(chip_yn == 1) |>
  mutate(
    # time to event 
    time = as.numeric((cancer_date_blood_leukemia_aml - dov)/365.25),
    end_date = case_when(
      # those who develop AML before cancer or death, among those who did not have previous cancer   
      cancer_dx_blood_leukemia_aml == 1 & prev_cancer_yn == 0 & 
        (cancer_date_blood_leukemia_aml <= cancer_date_cancer & cancer_date_blood_leukemia_aml <= date_of_death) ~ 
        cancer_date_blood_leukemia_aml, 
      # those who develop AML before cancer or death, among those who had previous cancer   
      cancer_dx_blood_leukemia_aml == 1 & prev_cancer_yn == 1 & 
        (cancer_date_blood_leukemia_aml > cancer_date_cancer & cancer_date_blood_leukemia_aml <= date_of_death) ~ 
        cancer_date_blood_leukemia_aml, 
      # those who develop cancer before developing AML or death 
      prev_cancer_yn == 0 & cancer_dx_cancer == 1 & 
        (cancer_date_cancer < cancer_date_blood_leukemia_aml & cancer_date_cancer < date_of_death) ~ cancer_date_cancer,
      # those who die without developing AML or cancer
      death == 1 & 
        (date_of_death < cancer_date_cancer | date_of_death < cancer_date_blood_leukemia_aml) ~ date_of_death, 
      # otherwise, administrative censoring  
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25, 
    
    high_risk_fct = droplevels(high_risk_fct),
    single_dnmt3a_fct = droplevels(single_dnmt3a_fct),
    mutation_n_fct = droplevels(mutation_n_fct), 
    max_vaf_fct = droplevels(max_vaf_fct), 
    chrs_cat = droplevels(chrs_cat),
    
    # MSS category, reduced
    mss_cat3 = case_when(
      mMSS < 0.51 ~ 0, 
      mMSS >= 0.51 & mMSS < 1.52 ~ 1, 
      mMSS >= 1.52 & !is.na(mMSS) ~ 2
    ),
    mss_cat3 = factor(
      mss_cat3,
      levels = c(0:2), 
      labels = c("MSS <0.51", "MSS 0.51-1.52", "MSS 1.52+")
    ),
    
    # raw scores + heteroplasmy yes/no, MSS in 3 categories for CHIP (+)    
    chrsm = (chrs + I(het_yn == 0)*1 + I(het_yn == 1)*1.5 + 
                  I(mss_cat3 == "MSS <0.51")*1 + I(mss_cat3 == "MSS 0.51-1.52")*1.5 + I(mss_cat3 == "MSS 1.52+")*2),
    chrsm = as.numeric(chrsm),
    
    # adding cutoff by 1 from original CHRS  
    chrsm_cat = case_when(
      chrsm <= 11.5 ~ 0,  
      chrsm >= 12 & chrsm <= 14 ~ 1, 
      chrsm >= 14.5 ~ 2
    ), 
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c(0:2), 
      labels = c("low", "intermediate", "high")
    )
  )

surv_chip <-
  with(ukb_chip, Surv(time = tt, event = cancer_dx_blood_leukemia_aml))

# CHRS-M category
chrsm_cat_aml <- survival::coxph(surv_chip ~ chrsm_cat + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chrsm_cat_aml)

# components of CHRS-M
chip_aml <- survival::coxph(surv_chip ~ 
                              age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct +
                              het_yn + mss_cat3 + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chip_aml)
```

## MDS  

```{r analysis figure 6 mds}
ukb_chip <- 
  ukb |>
  filter(chip_yn == 1) |>
  mutate(
    # time to event 
    time = as.numeric((cancer_date_unknown_mds - dov)/365.25),
    end_date = case_when(
      # those who develop MDS before cancer or death, among those who did not have previous cancer   
      cancer_dx_unknown_mds == 1 & prev_cancer_yn == 0 & 
        (cancer_date_unknown_mds <= cancer_date_cancer & cancer_date_unknown_mds <= date_of_death) ~ cancer_date_unknown_mds, 
      # those who develop MDS before cancer or death, among those who had previous cancer   
      cancer_dx_unknown_mds == 1 & prev_cancer_yn == 1 & 
        (cancer_date_unknown_mds > cancer_date_cancer & cancer_date_unknown_mds <= date_of_death) ~ cancer_date_unknown_mds, 
      # those who develop cancer before developing MDS or death 
      prev_cancer_yn == 0 & cancer_dx_cancer == 1 & 
        (cancer_date_cancer < cancer_date_unknown_mds & cancer_date_cancer < date_of_death) ~ cancer_date_cancer,
      # those who die without developing MDS or cancer
      death == 1 & 
        (date_of_death < cancer_date_cancer | date_of_death < cancer_date_unknown_mds) ~ date_of_death, 
      # otherwise, administrative censoring  
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25, 
    
    high_risk_fct = droplevels(high_risk_fct),
    single_dnmt3a_fct = droplevels(single_dnmt3a_fct),
    mutation_n_fct = droplevels(mutation_n_fct), 
    max_vaf_fct = droplevels(max_vaf_fct), 
    chrs_cat = droplevels(chrs_cat),
    
    # MSS category, reduced
    mss_cat3 = case_when(
      mMSS < 0.51 ~ 0, 
      mMSS >= 0.51 & mMSS < 1.52 ~ 1, 
      mMSS >= 1.52 & !is.na(mMSS) ~ 2
    ),
    mss_cat3 = factor(
      mss_cat3,
      levels = c(0:2), 
      labels = c("MSS <0.51", "MSS 0.51-1.52", "MSS 1.52+")
    ),
    
    # raw scores + heteroplasmy yes/no, MSS in 3 categories for CHIP (+)    
    chrsm = (chrs + I(het_yn == 0)*1 + I(het_yn == 1)*1.5 + 
                  I(mss_cat3 == "MSS <0.51")*1 + I(mss_cat3 == "MSS 0.51-1.52")*1.5 + I(mss_cat3 == "MSS 1.52+")*2),
    chrsm = as.numeric(chrsm),
    
    # adding cutoff by 1 from original CHRS  
    chrsm_cat = case_when(
      chrsm <= 11.5 ~ 0,  
      chrsm >= 12 & chrsm <= 14 ~ 1, 
      chrsm >= 14.5 ~ 2
    ), 
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c(0:2), 
      labels = c("low", "intermediate", "high")
    )
  )

surv_chip <-
  with(ukb_chip, Surv(time = tt, event = cancer_dx_unknown_mds))

# CHRS-M category
chrsm_cat_mds <- survival::coxph(surv_chip ~ chrsm_cat + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chrsm_cat_mds)

# components of CHRS-M
chip_mds <- survival::coxph(surv_chip ~ 
                              age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct +
                              het_yn + mss_cat3 + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chip_mds)
```

## MPN  

```{r analysis figure 6 mpn}
ukb_chip <- 
  ukb |>
  filter(chip_yn == 1) |>
  mutate(
    # time to event 
    time = as.numeric((cancer_date_blood_mpn - dov)/365.25),
    end_date = case_when(
      # those who develop MPN before cancer or death, among those who did not have previous cancer   
      cancer_dx_blood_mpn == 1 & prev_cancer_yn == 0 & 
        (cancer_date_blood_mpn <= cancer_date_cancer & cancer_date_blood_mpn <= date_of_death) ~ 
        cancer_date_blood_mpn, 
      # those who develop MPN before cancer or death, among those who had previous cancer   
      cancer_dx_blood_mpn == 1 & prev_cancer_yn == 1 & 
        (cancer_date_blood_mpn > cancer_date_cancer & cancer_date_blood_mpn <= date_of_death) ~ 
        cancer_date_blood_mpn, 
      # those who develop cancer before developing MPN or death 
      prev_cancer_yn == 0 & cancer_dx_cancer == 1 & 
        (cancer_date_cancer < cancer_date_blood_mpn & cancer_date_cancer < date_of_death) ~ cancer_date_cancer,
      # those who die without developing MPN or cancer
      death == 1 & 
        (date_of_death < cancer_date_cancer | date_of_death < cancer_date_blood_mpn) ~ date_of_death, 
      # otherwise, administrative censoring  
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25, 
    
    high_risk_fct = droplevels(high_risk_fct),
    single_dnmt3a_fct = droplevels(single_dnmt3a_fct),
    mutation_n_fct = droplevels(mutation_n_fct), 
    max_vaf_fct = droplevels(max_vaf_fct), 
    chrs_cat = droplevels(chrs_cat),
    
    # MSS category, reduced
    mss_cat3 = case_when(
      mMSS < 0.51 ~ 0, 
      mMSS >= 0.51 & mMSS < 1.52 ~ 1, 
      mMSS >= 1.52 & !is.na(mMSS) ~ 2
    ),
    mss_cat3 = factor(
      mss_cat3,
      levels = c(0:2), 
      labels = c("MSS <0.51", "MSS 0.51-1.52", "MSS 1.52+")
    ),
    
    # raw scores + heteroplasmy yes/no, MSS in 3 categories for CHIP (+)    
    chrsm = (chrs + I(het_yn == 0)*1 + I(het_yn == 1)*1.5 + 
                  I(mss_cat3 == "MSS <0.51")*1 + I(mss_cat3 == "MSS 0.51-1.52")*1.5 + I(mss_cat3 == "MSS 1.52+")*2),
    chrsm = as.numeric(chrsm),
    
    # adding cutoff by 1 from original CHRS  
    chrsm_cat = case_when(
      chrsm <= 11.5 ~ 0,  
      chrsm >= 12 & chrsm <= 14 ~ 1, 
      chrsm >= 14.5 ~ 2
    ), 
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c(0:2), 
      labels = c("low", "intermediate", "high")
    )
  )

surv_chip <-
  with(ukb_chip, Surv(time = tt, event = cancer_dx_blood_mpn))

# CHRS-M category
chrsm_cat_mpn <- survival::coxph(surv_chip ~ chrsm_cat + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chrsm_cat_mpn)

# components of CHRS-M
chip_mpn <- survival::coxph(surv_chip ~ 
                              age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct +
                              het_yn + mss_cat3 + sex + smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(chip_mpn)
```

## Combined  

```{r analysis figure 6, fig.height = 3, fig.width = 6}
table_mn <- 
  summary(chrsm_cat_mn)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mn") |>
  rownames_to_column("variable")

table_mds <- 
  summary(chrsm_cat_mds)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mds") |>
  rownames_to_column("variable")
 
table_aml <- 
  summary(chrsm_cat_aml)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "aml") |>
  rownames_to_column("variable")

table_mpn <- 
  summary(chrsm_cat_mpn)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mpn") |>
  rownames_to_column("variable")

table_chrsm <- 
  bind_rows(table_mn, table_aml, table_mds, table_mpn) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("mn", "mpn", "mds", "aml"),
      labels = c("MN", "MPN", "MDS", "AML")
    ),
    variable = factor(
      variable, 
      levels = c("chrsm_cathigh", "chrsm_catintermediate")
    )
  ) 

table_chrsm

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Figure 6", x = table_chrsm)

ggforestplot::forestplot(
  df = table_chrsm, 
  name = subtype, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = variable, 
  xlab = "HR (95% CI)"
  ) +
  scale_color_manual(
    "",
    values = c("#DF4375", "#027DFF" ),
    labels = c("CHRS-M: High risk", "CHRS-M: Intermediate risk")
  )
```

# Supplementary Figure 12 - Hazard ratios (95% confidence intervals) for each component of CHRS-M and subtypes of MN.      

```{r analysis suppl figure 12, fig.height = 6, fig.width = 8}
table_mn <- 
  summary(chip_mn)$coefficients[c(1:11), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mn") |>
  rownames_to_column("variable")

table_mds <- 
  summary(chip_mds)$coefficients[c(1:11), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mds") |>
  rownames_to_column("variable")

table_aml <- 
  summary(chip_aml)$coefficients[c(1:11), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "aml") |>
  rownames_to_column("variable")

table_mpn <- 
  summary(chip_mpn)$coefficients[c(1:11), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "mpn") |>
  rownames_to_column("variable")
 
table_components <- 
  bind_rows(table_mn, table_mds, table_aml, table_mpn) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("mn", "mpn", "mds", "aml"), 
      labels = c("MN", "MPN", "MDS", "AML")
    ),
    variable = factor(
      variable, 
      levels = c("age65_yn", "cytopenia_yn", "rdw_cat_yn", "mcv_cat_yn", "high_risk_fctyes", "single_dnmt3a_fctyes", "mutation_n_fct2+", "max_vaf_fct20+",
                 "het_yn", "mss_cat3MSS 0.51-1.52", "mss_cat3MSS 1.52+"),
      labels = c("Age 65+", "CCUS", "RDW 15+", "MCV 100+", "High-risk mutation", "Single DNMT3A", "Number of mutations 2+", "VAF 0.2+",
                 "Heteroplasmy", "mMSS 0.51-1.52", "mMSS 1.52+")
    )
  )

table_components 

ggforestplot::forestplot(
  df = table_components, 
  name = variable, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = subtype, 
  xlab = "HR (95% CI) for each variable"
  ) +
  scale_color_manual(
    "Subtypes", 
    values = c("black", "#74E1D5", "#104176", "#DF4375"),
    labels = c("MN", "MPN", "MDS", "AML")
  )

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Supplementary Figure 12", x = table_components)

# Export the file
openxlsx::saveWorkbook(wb, "ukb_mthet_chip_source_data_file_04_subgroup.xlsx", overwrite = TRUE)
```

