---
title: "UKB - Heteroplasmy, CHIP, and MN"
subtitle: "03 - prediction model for CHRS-M: Figure 5, Supplementary Figures 11, 16"  
author: "Yun Soo Hong"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false 
      smooth_scroll: false 
    pandoc_args: [
    "--number-sections",
    "--number-offset=0"
    ]
fontsize: 11pt              
---

# Set up  

## Program description  
* This program reads in the cleaned data set (`ukb_mthet_chip_dp3_2024_03_13.rds`), performs analysis in the "Mitochondrial heteroplasmy improves risk prediction for myeloid neoplasms" paper (Nature Communications, 2024), and provided the data used for generating tables and figures. May need ~200G to run comparison of ROC and bootstrapping.     
* This file contains
  -   Figure 5, Supplementary Figures 11, 16.    

* Steps  
  * Step 1: Read in data  
  * Step 2: Settings    
  * Step 3: Figure 5  
  * Step 4: Supplementary Table 3  
  * Step 5: Supplementary Figure 11  
  * Step 6: Supplementary Figure 16  
  * Step 7: Testing model performance  

## Version of R and packages  

```{r analysis set up, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries  

```{r analysis load libraries, echo = TRUE, message = FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(data.table)
library(survival)
library(survminer)
```

## Print out settings  

```{r analysis print settings}
options(tibble.print_max = 100)
options(pillar.sigfig = 5)
```

# Load data  

```{r analysis load data, include = FALSE}
ukb <- 
  readRDS("../data_cleaning/ukb_mthet_chip_dp3_2024_03_13.rds")
```

# Settings   

## Excel output  

```{r analysis excel output}
wb <- openxlsx::createWorkbook("")

# Add some sheets to the workbook
openxlsx::addWorksheet(wb, "Figure 5")
openxlsx::addWorksheet(wb, "Supplementary Figure 11")
openxlsx::addWorksheet(wb, "Supplementary Figure 16")
```

## Survival analysis    

### Time to event      

```{r analysis tte}
summary(ukb$cancer_date_blood_mn)

# all participants  
ukb <-
  ukb |>
  mutate(
    time = as.numeric((cancer_date_blood_mn - dov)/365.25),
    end_date = case_when(
      # those who develop MN before death   
      cancer_dx_blood_mn == 1 & 
        (cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      # those who die without developing MN
      death == 1 & 
        (date_of_death < cancer_date_blood_mn) ~ date_of_death, 
      # otherwise, administrative censoring  
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25
  )
summary(ukb$tt)

# CHIP positive  
ukb_chip <- 
  ukb |>
  filter(chip_yn == 1) |>
  mutate(
    high_risk_fct = droplevels(high_risk_fct),
    single_dnmt3a_fct = droplevels(single_dnmt3a_fct),
    mutation_n_fct = droplevels(mutation_n_fct), 
    max_vaf_fct = droplevels(max_vaf_fct), 
    chrs_cat = droplevels(chrs_cat),
    
    # MSS category, reduced
    mss_cat3 = case_when(
      mMSS < 0.51 ~ 0, 
      mMSS >= 0.51 & mMSS < 1.52 ~ 1, 
      mMSS >= 1.52 & !is.na(mMSS) ~ 2
    ),
    mss_cat3 = factor(
      mss_cat3,
      levels = c(0:2), 
      labels = c("MSS <0.51", "MSS 0.51-1.52", "MSS 1.52+")
    ),
    
    # raw scores + heteroplasmy yes/no, MSS in 3 categories for CHIP (+)    
    chrsm = (chrs + I(het_yn == 0)*1 + I(het_yn == 1)*1.5 + 
                  I(mss_cat3 == "MSS <0.51")*1 + I(mss_cat3 == "MSS 0.51-1.52")*1.5 + I(mss_cat3 == "MSS 1.52+")*2),
    chrsm = as.numeric(chrsm),
    
    # adding cutoff by 1 from original CHRS  
    chrsm_cat = case_when(
      chrsm <= 11.5 ~ 0,  
      chrsm >= 12 & chrsm <= 14 ~ 1, 
      chrsm >= 14.5 ~ 2
    ), 
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c(0:2), 
      labels = c("low", "intermediate", "high")
    ), 
    
    # incorporating reclassification 3 categories    
    chrs_het_recat3 = case_when(
      chrs_cat == "low" & chrsm_cat == "low" ~ 0,
      chrs_cat == "low" & chrsm_cat == "intermediate" ~ 1,
      chrs_cat == "intermediate" & chrsm_cat == "intermediate" ~ 2,
      chrs_cat == "intermediate" & chrsm_cat == "high" ~ 3,
      chrs_cat == "high" & chrsm_cat == "high" ~ 4
    ),
    chrs_het_recat3 = factor(
      chrs_het_recat3, 
      levels = c(0:4), 
      labels = c("low/low", "low/intermediate", "intermediate/intermediate", "intermediate/high", "high/high")
    )
  )
```

### Survival object        

```{r analysis obj}
surv <-
  with(ukb_chip, Surv(time = tt, event = cancer_dx_blood_mn))
```

# Figure 5 - Distribution of heteroplasmy and risk of MN by CHRS-M categories.  

## Panel A  

```{r analysis figure 5 a}
count <- 
  ukb_chip |>
  mutate(
    count_het_cat = forcats::fct_rev(count_het_cat)
  ) |> 
  ggplot() +  
  geom_bar(aes(x = chrsm_cat, fill = count_het_cat), position = "fill") +
  scale_x_discrete(
    "CHRS-M category", 
    labels=c("low" = "Low", "intermediate" = "Intermediate", "high" = "High")
  ) +
  scale_y_continuous(
    "Proportion", 
    breaks = c(0, 0.25, 0.5, 0.75, 1), 
    labels = c("0", "0.25", "0.50", "0.75", "1.0"), 
    limits = c(0, 1.01), 
    expand = c(0, 0)
  ) + 
  scale_fill_brewer("Heteroplasmy count", labels = c("4+", "3", "2", "1", "0"), palette = "Set1") +
  coord_flip() +
  theme_classic() + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5, nrow = 1, reverse = TRUE))
count

count_table <- 
  ukb_chip |>
  filter(chip_yn == 1) |>
  janitor::tabyl(count_het_cat, chrs_cat) |>
  janitor::adorn_percentages() |>
  janitor::adorn_ns() |>
  rename(
    Metric = count_het_cat
  )
```

## Panel B  

```{r analysis figure 5 b}
mss <- 
  ukb_chip |>
  filter(chip_yn == 1) |>
  ggplot() +
  geom_boxplot(aes(x = chrsm_cat, y = mMSS, fill = chrsm_cat)) +
  coord_flip() +
  geom_jitter(aes(x = chrsm_cat, y = mMSS, color = chrsm_cat), size = 0.1, width = 0.05, alpha = 0.5) +
  scale_x_discrete("", labels=c("low" = "Low", "intermediate" = "Intermediate", "high" = "High")) +
  scale_y_continuous("mMSS") + 
  scale_color_manual(
    "CHRS-M category", 
    labels = c("Low", "Intermediate", "High"), 
    values = c("#00A087FF", "#191970", "#E64835FF")
  ) + 
  scale_fill_manual(
    "CHRS-M category", 
    labels = c("Low", "Intermediate", "High"), 
    values = c("#00A087FF", "#191970", "#E64835FF")
  ) + 
  guides(color = "none") +
  theme_classic() + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  guides(fill = guide_legend(title.position = "top", title.hjust = 0.5, nrow = 1))
mss

mss_table <- 
  ukb_chip |>
  group_by(chrs_cat) |>
  summarise(
    min = min(mMSS), 
    q1 = quantile(mMSS, 0.25),
    median = median(mMSS), 
    q3 = quantile(mMSS, 0.75), 
    max = max(mMSS)
  ) |>
  t() |>
  as.data.frame() |>
  rownames_to_column() |>
  rename(
    Metric = rowname, 
    low = V1, 
    intermediate = V2, 
    high = V3
  ) |>
  filter(Metric != "chrs_cat")

table <- 
  count_table |>
  bind_rows(mss_table) |>
  add_row(.before = 1) |>
  add_row(.before = 7) |>
  mutate(
    Metric = c("Heteroplasmy count", "0", "1", "2", "3", "4+", "mMSS", "Minimum", "Q2", "Median", "Q3", "Maximum")
  )

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Figure 5", x = table)
```

## Panel C  

```{r analysis figure 5 c}
temp <- 
  ukb_chip |>
  mutate(
    chrsm_cat = factor(
      chrsm_cat, 
      levels = c("low", "intermediate", "high"),
      labels = c("low_m", "intermediate_m", "high_m")
    )
  ) |> 
  ggsankey::make_long(chrs_cat, chrsm_cat)

tally <-
  temp |>
  group_by(node) |>
  tally()

df_plot <- merge(temp, tally, by.x = 'node', by.y = 'node', all.x = TRUE)

plot_a_fit <- 
  df_plot |>
  mutate(
    x = factor(
      x,
      levels = c("chrs_cat", "chrsm_cat"),
      labels = c("CHRS", "CHRS-M")
    ),
    next_x = factor(
      next_x,
      levels = c("chrs_cat", "chrsm_cat"),
      labels = c("CHRS", "CHRS-M")
    ),
    node = factor(
      node,
      levels = c("low", "intermediate", "high", "low_m", "intermediate_m", "high_m"),
      labels = c("Low", "Intermediate", "High", "Low", "Intermediate", "High")
    ),
    next_node = factor(
      next_node,
      levels = c("low", "intermediate", "high", "low_m", "intermediate_m", "high_m"),
      labels = c("Low", "Intermediate", "High", "Low", "Intermediate", "High")
    )
  ) |> 
  ggplot(
    aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node), label = paste0(node, " N = ", n))) +
  ggsankey::geom_sankey(
    flow.alpha = 0.4
  ) +
  scale_fill_manual(
    "CHRS category",
    values = c("#00A087FF", "#191970", "#E64835FF")
  ) + 
  theme_minimal(base_size = 16) + 
  theme(legend.position = "none") + 
  ggsankey::geom_sankey_label(size = 3.5, color = "white", hjust = 0.2) +  
  theme(axis.title = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks = element_blank(), 
        panel.grid = element_blank())

combined_abc <- cowplot::plot_grid(count, mss, plot_a_fit, ncol = 3, rel_widths = c(1, 1, 1), labels = c("A", "B", "C"), label_x = c(0.1, 0.1, 0.1))
```

## Panel D     

```{r analysis figure 5 d}
km <- survfit(surv ~ chrs_het_recat3, data = ukb_chip)
1 - summary(km, times = 10)$surv

# KM plot  
s_km <- 
  summary(
    km,
    times = seq(0, 15, 1/365), 
    extend = TRUE
  ) 

plot_d_km_data <-
  tibble(
    time = s_km$time, 
    n_risk = s_km$n.risk, 
    n_event = s_km$n.event, 
    n_censor = s_km$n.censor, 
    surv = s_km$surv,
    cuminc = 1 - surv,
    strata = s_km$strata, 
    std_error = s_km$std.err, 
    ll95ci = 1 - s_km$upper, 
    ul95ci = 1 - s_km$lower
  )

plot_d_km <- 
  plot_d_km_data |>
  as.data.frame() |>
  ggplot(
    aes(x = time)
  ) +
  geom_step(
    aes(y = cuminc, color = strata, linetype = strata, group = strata), linewidth = 1
  ) +
  theme_classic(
    base_family = "sans", 
    base_size = 15
  ) +
  ggthemes::geom_rangeframe(
    data = data.frame(
      x = c(0, 15), 
      y = c(0, 0.5)
    ),
    aes(x, y), 
    color = c("black"), 
    size = 2
  ) + 
  scale_y_continuous(
    name = "Cumulative incidence (%)", 
    breaks = seq(0, 0.5, 0.1), 
    labels = c("0%", "10%", "20%", "30%", "40%", "50%")
  ) +
  scale_x_continuous(
    name = "Time since study enrollment (years)", 
    breaks = seq(0, 15, 5)
  ) + 
  scale_linetype_manual(name = "", values = c("twodash", "solid", "twodash", "solid", "twodash")) + 
  scale_color_manual(
    name = "",
    values = c("#00A087FF", "#00A087FD", "#191970", "#191971", "#E64835FF"),
    labels = c("CHRS: Low, CHRS-M: Low", "CHRS: Low, CHRS-M: Intermediate", "CHRS: Intermediate, CHRS-M: Intermediate", 
               "CHRS: Intermediate, CHRS-M: High", "CHRS: High, CHRS-M: High")
  ) +
  theme(
    axis.line = element_blank(),
    axis.title = element_text(face = 'bold'),
    panel.grid = element_blank(), 
    legend.position = "top",
    legend.key.spacing.x = unit(16, "pt"),
    legend.key.spacing.y = unit(1, "pt"),
    legend.margin = margin(t = 0, unit = "cm"), 
    legend.text = element_text(size = 10, margin = margin(l = 0))
  ) +
  guides(
    color = guide_legend(nrow = 3, byrow = TRUE, override.aes = list(linetype = c("twodash", "solid", "twodash", "solid", "twodash"))),
    linetype = "none"
  )

# Risk table  
s_table <- 
  summary(
    km,
    times = seq(0, 15, 5), 
    extend = TRUE
  ) 

plot_d_table_data <-
  tibble(
    time = s_table$time, 
    n_risk = s_table$n.risk, 
    n_event = s_table$n.event, 
    n_censor = s_table$n.censor, 
    surv = s_table$surv,
    cuminc = 1 - surv,
    strata = s_table$strata, 
    std_error = s_table$std.err, 
    ll95ci = 1 - s_table$upper, 
    ul95ci = 1 - s_table$lower
  )

plot_d_rt <-
  plot_d_table_data |>
  mutate(strata = factor(strata, levels = rev(levels(strata)))) |> 
  ggplot(
    aes(x = time, y = strata)
  ) +
  geom_text(
    aes(label = n_risk)
  ) + 
  ggthemes::geom_rangeframe(
    data = data.frame(
      x = c(0, 15), 
      y = c(1, 2)
    ),
    aes(x, y), 
    color = c("white"), 
    size = 2
  ) +
  scale_y_discrete(
    label = c("High -> High", "Int -> High", "Int -> Int", "Low -> Int", "Low -> Low")
  ) + 
  theme(
    axis.line = element_blank(), 
    axis.text.x = element_blank(), 
    axis.ticks = element_blank(), 
    axis.title = element_blank(), 
    panel.background = element_blank(), 
    panel.grid = element_blank()
  )

plot_d <- cowplot::plot_grid(plot_d_km, plot_d_rt, ncol = 1, rel_widths = c(1, 1), rel_heights = c(1, 0.25))
```

## Panel E     

```{r analysis figure 5 e}
km <- survfit(surv ~ chrs_het_recat3, data = ukb_chip, subset = chrs_het_recat3 == "low/low" | chrs_het_recat3 == "low/intermediate")

# KM curve  
s_km <- 
  summary(
    km,
    times = seq(0, 15, 1/365), 
    extend = TRUE
  ) 

plot_e_km_data <-
  tibble(
    time = s_km$time, 
    n_risk = s_km$n.risk, 
    n_event = s_km$n.event, 
    n_censor = s_km$n.censor, 
    surv = s_km$surv,
    cuminc = 1 - surv,
    strata = s_km$strata, 
    std_error = s_km$std.err, 
    ll95ci = 1 - s_km$upper, 
    ul95ci = 1 - s_km$lower
  )

plot_e_km <- 
  plot_e_km_data |>
  as.data.frame() |>
  ggplot(
    aes(x = time)
  ) +
  geom_step(
    aes(y = cuminc, color = strata, linetype = strata, group = strata), linewidth = 1
  ) +
  theme_classic(
    base_family = "sans", 
    base_size = 15
  ) +
  ggthemes::geom_rangeframe(
    data = data.frame(
      x = c(0, 15), 
      y = c(0, 0.02)
    ),
    aes(x, y), 
    color = c("black"), 
    size = 2
  ) + 
  scale_y_continuous(
    name = "Cumulative incidence (%)", 
    breaks = seq(0, 0.02, 0.005), 
    labels = c("0.0%", "0.5%", "1.0%", "1.5%", "2.0%")
  ) +
  scale_x_continuous(
    name = "Time since study enrollment (years)", 
    breaks = seq(0, 15, 5)
  ) + 
  scale_linetype_manual(name = "", values = c("twodash", "solid")) + 
  scale_color_manual(
    name = "",
    values = c("#00A087FF", "#00A087FD"),
    labels = c("CHRS: Low, CHRS-M: Low", "CHRS: Low, CHRS-M: Intermediate")
  ) +
  theme(
    axis.line = element_blank(),
    axis.title = element_text(face = 'bold'),
    panel.grid = element_blank(), 
    legend.position = "top", 
    legend.key.spacing.x = unit(12, "pt"),
    legend.key.spacing.y = unit(1, "pt"),
    legend.margin = margin(t = 0, unit = "cm"), 
    legend.text = element_text(size = 10, margin = margin(l = 0))
  ) +
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(linetype = c("twodash", "solid"))),
    linetype = "none"
  )

# Risk table  
s_table <- 
  summary(
    km,
    times = seq(0, 15, 5), 
    extend = TRUE
  ) 

plot_e_table_data <-
  tibble(
    time = s_table$time, 
    n_risk = s_table$n.risk, 
    n_event = s_table$n.event, 
    n_censor = s_table$n.censor, 
    surv = s_table$surv,
    cuminc = 1 - surv,
    strata = s_table$strata, 
    std_error = s_table$std.err, 
    ll95ci = 1 - s_table$upper, 
    ul95ci = 1 - s_table$lower
  )

plot_e_rt <-
  plot_e_table_data |>
  mutate(strata = factor(strata, levels = rev(levels(strata)))) |> 
  ggplot(
    aes(x = time, y = strata)
  ) +
  geom_text(
    aes(label = n_risk)
  ) + 
  ggthemes::geom_rangeframe(
    data = data.frame(
      x = c(0, 15), 
      y = c(1, 2)
    ),
    aes(x, y), 
    color = c("white"), 
    size = 2
  ) +
  scale_y_discrete(
    label = c("Low -> Int", "Low -> Low")
  ) + 
  theme(
    axis.line = element_blank(), 
    axis.text.x = element_blank(), 
    axis.ticks = element_blank(), 
    axis.title = element_blank(), 
    panel.background = element_blank(), 
    panel.grid = element_blank()
  )

plot_e <- cowplot::plot_grid(plot_e_km, plot_e_rt, ncol = 1, rel_widths = c(1, 1), rel_heights = c(1, 0.25))

combined_de <- cowplot::plot_grid(NULL, plot_d, plot_e, nrow = 1, rel_widths = c(0.1, 1, 1), labels = c("", "D", "E"))
```

## Combined  

```{r analysis figure 5, fig.height = 10, fig.width = 16}
combined <- cowplot::plot_grid(combined_abc, NULL, combined_de, ncol = 1, rel_heights = c(1, 0.15, 1))
combined
```

# Supplementary Figure 11 - Distribution and components of CHRS-M in UKB. (A) Histogram of the distribution of CHRS-M.        

## Panel A    

```{r analysis suppl figure 11 a}
hist_chrsm <- ukb_chip |>
  ggplot() +
  geom_histogram(
    aes(x = chrsm), 
    binwidth = 0.5,
    boundary = 0.25, 
    size = 0.5) +
  geom_text(
    aes(x = chrsm, y = after_stat(count), label = after_stat(count)),
    stat = "count",
    vjust = -0.5,
    size = 3
  ) +   
  scale_x_continuous(
   "CHRS-M",  
    breaks = seq(9.5, 18.5, by = 0.5)
  ) +
  scale_y_continuous(
    "Count", 
    expand = c(0, 0),
    limits = c(0, 12500),
    breaks = seq(0, 12500, by = 2500)
  ) +
  theme_classic()

ukb_chip |>
  group_by(chrsm) |>
  summarise(n = n())
```

## Panel B    

```{r analysis suppl figure 11 b}
chrsm <- survival::coxph(
  surv ~ age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct +
    sex + smk_ever + prev_cancer_yn + het_yn + mss_cat3, data = ukb_chip) 
summary(chrsm)

# forest plot for the estimates  
coef_all <- summary(chrsm)$coef[c(1:8, 12:14), c(1:3, 5)]

table <- 
  coef_all |> 
  as_tibble() |> 
  mutate(
    var = row.names(coef_all), 
    var = case_when(
      var == "age65_yn" ~ "Age \u2265 65 years",
      var == "cytopenia_yn" ~ "CCUS", 
      var == "rdw_cat_yn" ~ "RDW \u2265 15", 
      var == "mcv_cat_yn" ~ "MCV \u2265 100", 
      var == "high_risk_fctyes" ~ "High-risk mutation", 
      var == "single_dnmt3a_fctyes" ~ "Single DNMT3A", 
      var == "mutation_n_fct2+" ~ "Number of mutations", 
      var == "max_vaf_fct20+" ~ "VAF \u2265 0.2",
      var == "het_yn" ~ "Presence of heteroplasmy",
      var == "mss_cat3MSS 0.51-1.52" ~ "MSS 0.51-1.52",
      var == "mss_cat3MSS 1.52+" ~ "MSS 1.52+"
    )
  ) |>
  rename(
    `P value` = `Pr(>|z|)`,
    Variable = var
  ) |> 
  mutate(
    # add estimates  
    lci = exp(coef - 1.96*`se(coef)`),
    uci = exp(coef + 1.96*`se(coef)`),
    `HR (95% CI)` = sprintf("%.2f (%.2f to %.2f)", `exp(coef)`, lci, uci),
    ` ` = paste(rep(" ", 25), collapse= " "),
    `P value` = format(`P value`, scientific = TRUE, digits = 3)
  ) |>
  dplyr::select(
    Variable, ` `, `HR (95% CI)`, everything()
  ) 
table

p_chrsm <- forestploter::forest(
  table[, c(1:3, 8)],
  est = table$`exp(coef)`,
  lower = table$lci,
  upper = table$uci,
  ci_column = 2,
  ref_line = 1,
  x_trans = "log",
  xlim = c(0.5, 8), 
  ticks_at = c(0.5, 1, 2, 4, 8)
  )
```

## Combined  

```{r analysis suppl figure 11, fig.height = 5, fig.width = 15}
combined <- ggpubr::ggarrange(hist_chrsm, p_chrsm, nrow = 1, labels = c("A", "B"), widths = c(0.8, 1))
combined

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Supplementary Figure 11", x = table)
```

# Supplementary Figure 16 - Linear association between mMSS and incident MN used to generate CHRS-M.  

```{r analysis suppl figure 16}
het_mss_linear <- 
  survival::coxph(surv ~ het_yn + mMSS + 
                    age65_yn + cytopenia_yn + rdw_cat_yn + mcv_cat_yn + high_risk_fct + single_dnmt3a_fct + mutation_n_fct + max_vaf_fct + sex +
                    smk_ever + prev_cancer_yn, data = ukb_chip) 
summary(het_mss_linear)

# linear model  
range_mss <- range(ukb_chip$mMSS)
range_mss_cut <- seq(range_mss[1], range_mss[2], length.out = 500) |>
  as.matrix()

beta <- coef(het_mss_linear)[2]
beta

var <- vcov(het_mss_linear)[2, 2]
var

loghr <- as.vector(range_mss_cut %*% beta)
loghrse <- sqrt(diag(range_mss_cut %*% var %*% t(range_mss_cut)))
loghr_ll <- as.vector(loghr - 1.96*loghrse)
loghr_ul <- as.vector(loghr + 1.96*loghrse)

hrplot <- 
  range_mss_cut |>
  as.data.frame() |> 
  rename(mMSS = V1) |> 
  cbind(loghr, loghrse, loghr_ll, loghr_ul) 

hrplot <-
  hrplot |>
  mutate(upper = exp(loghr_ul),
         lower = exp(loghr_ll),
         hr = exp(loghr))

p_loghr <- 
  hrplot |> 
  ggplot() +
  geom_ribbon(aes(x = mMSS, ymin = loghr_ll, ymax = loghr_ul), fill = "purple3", alpha = 0.3) +
  geom_line(aes(x = mMSS, y = loghr), linetype = "solid", size = 1.5, color = "purple3") +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             size = 1) +
  geom_hline(yintercept = 0.25,
             linetype = "dotted",
             color = "blue", 
             size = 0.5) +
  geom_hline(yintercept = 0.75,
             linetype = "dotted",
             color = "blue", 
             size = 0.5) +
  geom_hline(yintercept = 1.25,
             linetype = "dotted",
             color = "blue", 
             size = 0.5) +
  geom_vline(xintercept = 0.51,
             linetype = "dotted",
             color = "red", 
             size = 0.5) +
  geom_vline(xintercept = 1.52,
             linetype = "dotted",
             color = "red", 
             size = 0.5) +
  geom_vline(xintercept = 2.54,
             linetype = "dotted",
             color = "red", 
             size = 0.5) +
  scale_x_continuous(name = "mMSS",
                     expand = c(0.01, 0.01),
                     breaks = seq(0, 3, by = 0.5),
                     limits = c(0, 3)) +
  scale_y_continuous(name = "log(HR) for MN",
                     breaks = seq(0, 3, by = 0.5),
                     limits = c(-0.2, 3)) +
  theme_classic() +
  theme(axis.title = element_text(face = "bold",
                                  size = 14),
        axis.text = element_text(size = 12))
p_loghr

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Supplementary Figure 16", x = hrplot)

# Export the file
openxlsx::saveWorkbook(wb, "ukb_mthet_chip_source_data_file_03_prediction.xlsx", overwrite = TRUE)
```

# Testing model performance  

## NRI  

```{r analysis nri}
set.seed(20240702)

time = ukb_chip$tt
event = ukb_chip$cancer_dx_blood_mn

# using 3 categories  
z.std = as.matrix(subset(ukb_chip, select = chrs_cat))
z.new = as.matrix(subset(ukb_chip, select = chrsm_cat))

mstd = coxph(Surv(time, event) ~ ., data.frame(time, event, z.std), x = TRUE)
mnew = coxph(Surv(time, event) ~ ., data.frame(time, event, z.new), x = TRUE)

nricens::nricens(
  mdl.std = mstd, 
  mdl.new = mnew, 
  t0 = 10, 
  cut = c(0.02, 0.1), 
  niter = 100
)
```

## AUROC  

```{r analysis auroc}
# CHRS  
Hmisc::rcorrcens(
  Surv(tt, cancer_dx_blood_mn) ~ I(-1*as.numeric(chrs))  + I(-1*as.numeric(sex)) + I(-1*as.numeric(smk_ever)) + I(-1*prev_cancer_yn), 
  data = ukb_chip)

# CHRS-M  
Hmisc::rcorrcens(
  Surv(tt, cancer_dx_blood_mn) ~ I(-1*as.numeric(chrsm))  + I(-1*as.numeric(sex)) + I(-1*as.numeric(smk_ever)) + I(-1*prev_cancer_yn), 
  data = ukb_chip)

# CHRS-M categories  
Hmisc::rcorrcens(
  Surv(tt, cancer_dx_blood_mn) ~ I(-1*as.numeric(chrsm_cat))  + I(-1*as.numeric(sex)) + I(-1*as.numeric(smk_ever)) + I(-1*prev_cancer_yn), 
  data = ukb_chip)

# Comparing chrs with chrs_het  
roc_chrs <- 
  timeROC::timeROC(
    T = ukb_chip$tt, 
    delta = ukb_chip$cancer_dx_blood_mn, 
    marker = ukb_chip$chrs, 
    cause = 1, 
    weighting = "marginal", 
    times = c(5, 10),
    iid = TRUE
  )
roc_chrsm <- 
  timeROC::timeROC(
    T = ukb_chip$tt, 
    delta = ukb_chip$cancer_dx_blood_mn, 
    marker = ukb_chip$chrsm, 
    cause = 1, 
    weighting = "marginal", 
    times = c(5, 10),
    iid = TRUE
  )
roc_chrsm_cat <- 
  timeROC::timeROC(
    T = ukb_chip$tt, 
    delta = ukb_chip$cancer_dx_blood_mn, 
    marker = ukb_chip$chrsm_cat, 
    cause = 1, 
    weighting = "marginal", 
    times = c(5, 10),
    iid = TRUE
  )

# compare results  
roc_chrs 
confint(roc_chrs)

roc_chrsm
confint(roc_chrsm)

roc_chrsm_cat
confint(roc_chrsm_cat)

# p-values comparing the auroc  
timeROC::compare(roc_chrs, roc_chrsm) 
timeROC::compare(roc_chrs, roc_chrsm, adjusted=TRUE)
```

## Brier score  

```{r analysis brier}
set.seed(20240702)

# Set the number of bootstrap iterations
n_bootstrap <- 500

# Initialize vectors to store differences in Brier scores
diff_brier_scores <- numeric(n_bootstrap)

# Perform bootstrapping
for (i in 1:n_bootstrap) {
  # Generate a bootstrap sample
  bootstrap_sample <- ukb_chip[sample(nrow(ukb_chip), replace = TRUE), ]
  
  # Fit model1 and model2 to the bootstrap sample
  model1_bs <- survival::coxph(Surv(tt, cancer_dx_blood_mn) ~ chrs + sex + smk_ever + prev_cancer_yn, data = bootstrap_sample)
  model2_bs <- survival::coxph(Surv(tt, cancer_dx_blood_mn) ~ chrsm + sex + smk_ever + prev_cancer_yn, data = bootstrap_sample)
  
  # Calculate Brier scores for both models
  brier1 <- survival::brier(model1_bs, times = 10)$brier
  brier2 <- survival::brier(model2_bs, times = 10)$brier
  
  # Compute the difference in Brier scores
  diff_brier_scores[i] <- brier1 - brier2
}

head(diff_brier_scores, n = 10)

# Calculate the 95% CI
median <- quantile(diff_brier_scores, 0.50)
ci_lower <- quantile(diff_brier_scores, 0.025)
ci_upper <- quantile(diff_brier_scores, 0.975)

# Print the results
cat("Median, 95% CI for the difference in Brier scores:",
    round(median, 5), ", ", round(ci_lower, 5), "to", round(ci_upper, 5), "\n")
```





