---
title: "UKB - Heteroplasmy, CHIP, and MN"
subtitle: "05 - survival analysis by VAF: Supplementary Figure 15"  
author: "Yun Soo Hong"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false 
      smooth_scroll: false 
    pandoc_args: [
    "--number-sections",
    "--number-offset=0"
    ]
fontsize: 11pt              
---

# Set up  

## Program description  
* This program reads in the cleaned data set (`ukb_mthet_chip_dp3_2024_03_13.rds`), performs analysis in the "Mitochondrial heteroplasmy improves risk prediction for myeloid neoplasms" paper (Nature Communications, 2024), and provided the data used for generating tables and figures.   
* This file contains
  -   Supplementary Figure 15.    

* Steps  
  * Step 1: Read in data  
  * Step 2: Settings    
  * Step 3: Supplementary Figure 15    

## Version of R and packages  

```{r analysis set up, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries  

```{r analysis load libraries, echo = TRUE, message = FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(data.table)
library(survival)
library(survminer)
```

## Print out settings  

```{r analysis print settings}
options(tibble.print_max = 100)
options(pillar.sigfig = 5)
```

# Load data  

```{r analysis load data, include = FALSE}
vaf3 <- 
  readRDS("../data_cleaning/ukb_mthet_chip_vaf3_dp3_2024_03_13.rds")

vaf5 <- 
  readRDS("../data_cleaning/ukb_mthet_chip_dp3_2024_03_13.rds")

vaf10 <- 
  readRDS("../data_cleaning/ukb_mthet_chip_vaf10_dp3_2024_03_13.rds")
```

# Settings   

## Excel output  

```{r analysis excel output}
wb <- openxlsx::createWorkbook("")

# Add some sheets to the workbook
openxlsx::addWorksheet(wb, "Supplementary Figure 15")
```

## Survival analysis  

### Time to event  

```{r analysis tte}
vaf3 <-
  vaf3 |>
  mutate(
    end_date = case_when(
      cancer_dx_blood_mn == 1 & 
        (cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      death == 1 & 
        (date_of_death < cancer_date_blood_mn) ~ date_of_death, 
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25
  )

vaf5 <-
  vaf5 |>
  mutate(
    end_date = case_when(
      cancer_dx_blood_mn == 1 & 
        (cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      death == 1 & 
        (date_of_death < cancer_date_blood_mn) ~ date_of_death, 
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25
  )

vaf10 <-
  vaf10 |>
  mutate(
    end_date = case_when(
      cancer_dx_blood_mn == 1 & 
        (cancer_date_blood_mn <= date_of_death) ~ cancer_date_blood_mn, 
      death == 1 & 
        (date_of_death < cancer_date_blood_mn) ~ date_of_death, 
      TRUE ~ as.Date("2022-12-20")
    ), 
    tt = as.numeric(end_date - dov)/365.25
  ) 
```

### Survival object  

```{r analysis obj}
surv_vaf3 <-
  with(vaf3, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf5 <-
  with(vaf5, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf10 <-
  with(vaf10, Surv(time = tt, event = cancer_dx_blood_mn))
```

# Supplementary Figure 15 - Comparison of hazard ratios (95% confidence intervals) for the associations of heteroplasmy, CHIP, and incident MN by different VAF cut-offs for heteroplasmy.    

## Panel A  

```{r analysis suppl figure 15 a}
# VAF 3% 
vaf3_het_chip <- survival::coxph(surv_vaf3 ~ het_chip + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf3) 
summary(vaf3_het_chip)

table_vaf3 <- 
  summary(vaf3_het_chip)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf3") |>
  rownames_to_column("variable")
table_vaf3

# VAF 5%
vaf5_het_chip <- survival::coxph(surv_vaf5 ~ het_chip + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf5) 
summary(vaf5_het_chip)

table_vaf5 <- 
  summary(vaf5_het_chip)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf5") |>
  rownames_to_column("variable")
table_vaf5

# VAF 10% 
vaf10_het_chip <- survival::coxph(surv_vaf10 ~ het_chip + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf10) 
summary(vaf10_het_chip)

table_vaf10 <- 
  summary(vaf10_het_chip)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf10") |>
  rownames_to_column("variable")
table_vaf10

het_chip <- 
  bind_rows(table_vaf3, table_vaf5, table_vaf10) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("vaf10", "vaf5", "vaf3")
    ),
    variable = factor(
      variable, 
      levels = c("het_chiphet only", "het_chipchip only", "het_chipboth"),
      labels = c("Heteroplasmy only", "CHIP only", "Both")
    )
  ) |>
  mutate(
    Panel = "A"
  )
het_chip

p_het_chip <- 
  ggforestplot::forestplot(
  df = het_chip, 
  name = variable, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = subtype, 
  xlab = "HR (95% CI)"
  ) +
  scale_color_manual(
    "VAF cutoff", 
    values = c("#DF4375", "#104176", "#74E1D5"),
    labels = c("VAF 10%", "VAF 5%", "VAF 3%")
  ) 
```

## Panel B    

```{r analysis suppl figure 15 b}
# VAF 3%
vaf3_chrs <- survival::coxph(surv_vaf3 ~ chrs_cat + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf3) 
summary(vaf3_chrs)

table_vaf3 <- 
  summary(vaf3_chrs)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf3") |>
  rownames_to_column("variable")
table_vaf3

# VAF 5% 
vaf5_chrs <- survival::coxph(surv_vaf5 ~ chrs_cat + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf5) 
summary(vaf5_chrs)

table_vaf5 <- 
  summary(vaf5_chrs)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf5") |>
  rownames_to_column("variable")
table_vaf5

# VAF 10%  
vaf10_chrs <- survival::coxph(surv_vaf10 ~ chrs_cat + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf10) 
summary(vaf10_chrs)

table_vaf10 <- 
  summary(vaf10_chrs)$coefficients[c(1:3), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf10") |>
  rownames_to_column("variable")
table_vaf10

chrs <- 
  bind_rows(table_vaf3, table_vaf5, table_vaf10) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("vaf10", "vaf5", "vaf3")
    ),
    variable = factor(
      variable, 
      levels = c("chrs_catlow", "chrs_catintermediate", "chrs_cathigh"),
      labels = c("Low", "Intermediate", "High")
    )
  ) |>
  mutate(
    Panel = "B"
  )
chrs

p_chrs <- 
  ggforestplot::forestplot(
  df = chrs, 
  name = variable, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = subtype, 
  xlab = "HR (95% CI)"
  ) +
  scale_color_manual(
    "VAF cutoff", 
    values = c("#DF4375", "#104176", "#74E1D5"),
    labels = c("VAF 10%", "VAF 5%", "VAF 3%")
  ) 
```

## Panel C    

```{r analysis suppl figure 15 c}
# restriction to CHIP (+) individuals  
vaf3_chip <-
  vaf3 |>
  filter(chip_yn == 1) 

vaf5_chip <-
  vaf5 |>
  filter(chip_yn == 1) 

vaf10_chip <-
  vaf10 |>
  filter(chip_yn == 1) 

surv_vaf3 <-
  with(vaf3_chip, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf5 <-
  with(vaf5_chip, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf10 <-
  with(vaf10_chip, Surv(time = tt, event = cancer_dx_blood_mn))

# VAF 3% 
vaf3_het <- survival::coxph(surv_vaf3 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf3_chip) 
summary(vaf3_het)

table_vaf3 <- 
  summary(vaf3_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf3") |>
  rownames_to_column("variable")
table_vaf3

# VAF 5% 
vaf5_het <- survival::coxph(surv_vaf5 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf5_chip) 
summary(vaf5_het)

table_vaf5 <- 
  summary(vaf5_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf5") |>
  rownames_to_column("variable")
table_vaf5

# VAF 10%
vaf10_het <- survival::coxph(surv_vaf10 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf10_chip) 
summary(vaf10_het)

table_vaf10 <- 
  summary(vaf10_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf10") |>
  rownames_to_column("variable")
table_vaf10

het_cp <- 
  bind_rows(table_vaf3, table_vaf5, table_vaf10) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("vaf10", "vaf5", "vaf3")
    ),
    variable = factor(
      variable, 
      levels = c("count_het", "mMSS"),
      labels = c("Heteroplasmy count", "mMSS")
    )
  ) |>
  mutate(
    Panel = "C"
  )
het_cp

p_het_cp <- 
  ggforestplot::forestplot(
  df = het_cp, 
  name = variable, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = subtype, 
  xlab = "HR (95% CI)"
  ) +
  scale_color_manual(
    "VAF cutoff", 
    values = c("#DF4375", "#104176", "#74E1D5"),
    labels = c("VAF 10%", "VAF 5%", "VAF 3%")
  ) 
```

## Panel D    

```{r analysis suppl figure 15 d}
vaf3_nochip <-
  vaf3 |>
  filter(chip_yn == 0)

vaf5_nochip <-
  vaf5 |>
  filter(chip_yn == 0)

vaf10_nochip <-
  vaf10 |>
  filter(chip_yn == 0)

surv_vaf3 <-
  with(vaf3_nochip, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf5 <-
  with(vaf5_nochip, Surv(time = tt, event = cancer_dx_blood_mn))

surv_vaf10 <-
  with(vaf10_nochip, Surv(time = tt, event = cancer_dx_blood_mn))

# VAF 3%
vaf3_het <- survival::coxph(surv_vaf3 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf3_nochip) 
summary(vaf3_het)

table_vaf3 <- 
  summary(vaf3_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf3") |>
  rownames_to_column("variable")
table_vaf3

# VAF 5%
vaf5_het <- survival::coxph(surv_vaf5 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf5_nochip) 
summary(vaf5_het)

table_vaf5 <- 
  summary(vaf5_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf5") |>
  rownames_to_column("variable")
table_vaf5

# VAF 10%
vaf10_het <- survival::coxph(surv_vaf10 ~ count_het + mMSS + rms::rcs(age, 5) + sex + smk_ever + prev_cancer_yn, data = vaf10_nochip) 
summary(vaf10_het)

table_vaf10 <- 
  summary(vaf10_het)$coefficients[c(1:2), c(1:3, 5)] |>
  as.data.frame() |>
  mutate(subtype = "vaf10") |>
  rownames_to_column("variable")
table_vaf10

het_cn <- 
  bind_rows(table_vaf3, table_vaf5, table_vaf10) |>
  rename(
    se = `se(coef)`
  ) |>
  mutate(
    hr = exp(coef), 
    lci = exp(coef - 1.96*se), 
    uci = exp(coef + 1.96*se),
    subtype = factor(
      subtype, 
      levels = c("vaf10", "vaf5", "vaf3")
    ),
    variable = factor(
      variable, 
      levels = c("count_het", "mMSS"),
      labels = c("Heteroplasmy count", "mMSS")
    )
  ) |>
  mutate(
    Panel = "D"
  )
het_cn

p_het_cn <- 
  ggforestplot::forestplot(
  df = het_cn, 
  name = variable, 
  estimate = coef, 
  se = se, 
  logodds = TRUE, 
  colour = subtype, 
  xlab = "HR (95% CI)"
  ) +
  scale_color_manual(
    "VAF cutoff", 
    values = c("#DF4375", "#104176", "#74E1D5"),
    labels = c("VAF 10%", "VAF 5%", "VAF 3%")
  ) 
```

## Combined  

```{r analysis suppl figure 15, fig.height = 5, fig.width = 10}
combined <- ggpubr::ggarrange(
  p_het_chip, p_chrs, p_het_cp, p_het_cn,
  nrow = 2, ncol = 2,
  common.legend = TRUE,
  labels = c("A", "B", "C", "D")
)
combined

table <- 
  het_chip |>
  bind_rows(chrs) |> 
  bind_rows(het_cp) |>
  bind_rows(het_cn) |>
  dplyr::select(Panel, everything())
table

# Write the data to the sheets
openxlsx::writeData(wb, sheet = "Supplementary Figure 15", x = table)

# Export the file
openxlsx::saveWorkbook(wb, "ukb_mthet_chip_source_data_file_05_vaf.xlsx", overwrite = TRUE)
```

